---
title: "Class 09"
author: "Amy Prichard"
date: "February 8, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Unsupervised Learning Analysis of Human Breast Cancer Cells

## 1. Exploratory Data Analysis

```{r}
# Save your input data file to a new 'data' directory
fna.data <- "data/WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data)

head(wisc.df)
# looks like there is a funny last column called "X" --> need to get rid of it
```

Let's take columns 3-32 for further analysis (i.e. get rid of patient ID, diagnosis, and "column X" so they don't clog up the data)
```{r}
# Convert the features of the data: wisc.data
wisc.data <- as.matrix(wisc.df[,3:32])
```

...but we want to keep patient ID as ROW NAMES!
```{r}
# Set the row names of wisc.data
rownames(wisc.data) <- wisc.df$id

head(wisc.data)
```

...and we want to keep the diagnoses as INDECES!
```{r}
# Create diagnosis vector by completing the missing code
diagnosis <- as.numeric(wisc.df$diagnosis=="M")
```

### Q1. How many observations are in this dataset?
```{r}
# Check how many patients
nrow(wisc.data)  # 569

# Check how many features tested
ncol(wisc.data)  #  33
```

### Q2. How many variables/features in he data are suffixed with "_mean"?
```{r}
# Regular Expressions... or not.
# grep() finds indeces that match some pattern
# in this case, the pattern is "_mean" but it could also be a regular expression
# fixed=TRUE means the pattern must be a string that should be matched sa is
# fixed=FALSE is the default... I imagine it allows regular expressions?
length(grep("_mean", colnames(wisc.data), fixed=TRUE))  # fixed=TRUE seems optional?

# Python/R version
# meancounter = 0
# for item in colnames(wisc.data):
#     if "_mean" in item:
#         meancounter+=1
# print(meancounter)
```


### Q3. How many of the observations have a malignant diagnosis?
```{r}
table(wisc.df$diagnosis)
```

## Principal Component Analysis

```{r}
apply(wisc.data, 2, mean)  # 2 indicates columns (1 indicates rows)
```

```{r}
round(colMeans(wisc.data), 2)  # another way to do it

# what about standard deviation?
round(apply(wisc.data,2,sd), 2)
```

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale=TRUE)

# Look at summary of results
summary(wisc.pr)
```

### Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?
44.27%

### Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?
Three

### Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?
Seven

```{r}
biplot(wisc.pr)
```

### Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?
"That is a useless plot!"

```{r}
# Scatter plot observations by components 1 and 2
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=wisc.df$diagnosis, xlab="PC1", ylab="PC2")
```

### Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?
```{r}
# Repeat for components 1 and 3
plot(wisc.pr$x[,1], wisc.pr$x[,3], col=wisc.df$diagnosis, xlab="PC1", ylab="PC3")
```
The two graphs look pretty similar, but the PC1 vs PC3 graph is shifted a bit downward relative to the PC1 vs PC2 graph.

```{r}
# Making a scree plot...

# calculate variance
prvar <- wisc.pr$sdev^2

# calculate the variance explained by each prinicpal component
pve <- round((prvar/sum(prvar))*100, 2)

# plot it!
plot(pve, type="o", xlab="Principal Component", ylab="Proportion of Variance Explained", main="Scree Plot")
```

Fancy barplot
```{r}
barplot(pve, ylab = "Percent of Variance Explained", names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
# names.arg=paste0("PC",1:length(pve)) labels the bars on the barplot
          # paste0() sets sep="" to get rid of whitespace
# las=2 makes the axis labels perpendicular (see graphical parameters)
# axes=FALSE gets rid of existing axes
axis(2, at=pve, labels=round(pve,2)*100 )
# this puts a custom axis on the graph
# at=pve puts the ticks where the values are
# labels=round(pve,2)*100 rounds the tick labels
```



```{r}
# Plot cumulative proportion of variance explained
plot(cumsum(pve), xlab = "Principal Component", ylab = "Cumulative Proportion of Variance Explained", ylim = c(0, 1), type = "o")

# not sure why this is blank
```

```{r}
## ggplot based graph
# install.packages("factoextra")
library(factoextra)
```

```{r}
fviz_eig(wisc.pr, addlabels = TRUE)
```

### Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

### Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

## Hierarchical Clustering

```{r}
# Scale the wisc.data data: data.scaled
data.scaled <- prcomp(wisc.data, scale=TRUE)

data.dist <- dist(wisc.pr$x[,1:2])

wisc.hclust <- hclust(data.dist, method="ward.D2")
```

### Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?
```{r}
plot(wisc.hclust)
abline(h=36.7, col="red", lty=2)
wisc.hclust.clusters <- cutree(wisc.hclust, h=36.7)
table(wisc.hclust.clusters, diagnosis)  # cross tabulation
```

### Q12. Can you find a better cluster vs diagnoses match with by cutting into a different number of clusters between 2 and 10?
```{r}
table(cutree(wisc.hclust, k=2), diagnosis)  # makes logical sense because Malignant vs Benign tumors
```


We want 3 groups...
```{r}
grp3 <- cutree(wisc.hclust, k=3)
table(grp3)

plot(wisc.pr$x[,1], wisc.pr$x[,2], xlab="PC1", ylab="PC2", col=grp3)
```

Cross tabulation of the 3 groups
```{r}
table(grp3, wisc.df$diagnosis)
```


## K-means Clustering

```{r}
wisc.km <- kmeans(wisc.data, centers=2, nstart=20)
```

```{r}
table(wisc.km$cluster, diagnosis)
```

### Q13. How well does k-means separate the two diagnoses? How does it compare to your hclust results?
This seems better than hclust... but probably not perfect

```{r}
table(wisc.hclust.clusters, wisc.km$cluster)
```

## Combining methods

## Sensitivity/Specificity



# ~~~

## Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=wisc.df$diagnosis)
points(npc[,1], npc[,2], col=c("yellow","blue"), pch=15, cex=3)
```

### Q17. Which of these new patients should we prioritize for follow up based on your results?
The blue one (patient #2)