---
title: 'Class 7: Functions and Packages'
author: "Amy Prichard"
date: "January 30, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Functions revisited

Load (i.e. **source**) our rescale() function from last day.

```{r}
source("http://tinyurl.com/rescale-R")
# loads several functions and couple data frames into the global environment (df1, df2, df3, both_na, both_na2, both_na3, gene_intersect, gene_intersect2, gene_intersect3, gene_intersect4, rescale, rescale2)
```

Test this function

```{r}
rescale(1:5)
```

```{r}
rescale(c(1:5,"string"))
```

We want to make this function more robust to these types of errors

warning()  # Gives the user an error message without stopping the function.
stop()  # Gives the user an error and stops running the function.

rescale2() starts with
if(!is.numeric(x)) {stop("Input x should be numeric", call.=FALSE)}
to stop the function if the input is not numeric.

```{r}
rescale2(c(1:5,"string"))
```

```{r}
x <- c( 1, 2, NA, 3, NA)
y <- c(NA, 3, NA, 3,  4)

is.na(x)
is.na(y)
```

```{r}
is.na(x) & is.na(y)         # When are BOTH lists "NA"?
sum(is.na(x) & is.na(y))    # How many times does this happen?
which(is.na(x) & is.na(y))  # Where does this happen?
```

Now take our working snippet (sum(is.na(x) & is.na(y))) and make a function: "both_na()" which tells you how many overlapping "NA" values there are

But... what if the two lists/vectors have different numbers of items?

```{r}
x1 <- c(NA, NA, NA)
y1 <- c(1, NA, NA)

both_na(x1, y1)  # This is ok (as expected).
```

```{r}
y2 <- c(1, NA, NA, NA)
y3 <- c(1, NA, NA, NA, NA)

both_na(x1, y2)  # This doesn't work like we want it to.
```

Add a stop()

both_na2() starts with
if(length(x) != length(y)) {stop("Input x and y should be vectors of the same length", call.=FALSE)}

```{r}
both_na2(x1, y2)
```

Refine and polish: make our function more useful by returning more information *(see both_na3)*

```{r}
both_na3(x, y)
```

